Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-05-20 _tex/paper2.tex
Built site for gh-pages

Numerous studies have documented the effectiveness of CJ in assessing
traits and competencies over the past decade. These studies have
highlighted three aspects of the method's effectiveness: its
reliability, validity, and practical applicability. Research on
reliability suggests that CJ requires a relatively modest number of
pairwise comparisons \citep{Verhavert_et_al_2019, Crompvoets_et_al_2022}
to generate trait scores that are as precise and consistent as those
generated by other assessment methods
\citep{Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}. In
addition, the evidence suggests that the reliability and time efficiency
of CJ are comparable, if not superior, to those of other assessment
methods when employing adaptive comparison algorithms
\citep{Pollitt_2012b, Verhavert_et_al_2022, Mikhailiuk_et_al_2021}.
Meanwhile, research on validity indicates the capacity of CJ scores to
represent accurately the traits under measurement
\citep{Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Bartholomew_et_al_2018, Bouwer_et_al_2023}.
Lastly, research on its practical applicability highlights CJ's
versatility across both educational and non-educational contexts
\citep{Kimbell_2012, Jones_et_al_2015, Bartholomew_et_al_2018, Jones_et_al_2019, Marshall_et_al_2020, Bartholomew_et_al_2020, Boonen_et_al_2020}.

@@

In the first scenario, research on text quality suggests that when
judges evaluate multidimensional, complex traits or heterogeneous
stimuli, they often rely on various intricate characteristics of the
stimuli to form their judgments
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}. These
additional relevant characteristics, when assessed, are unlikely to be
equally weighted or opposing. As a result, they may exert an uneven
influence on judges' perceptions, creating biases that resist
cancellation. For example, this could occur when a judge assessing the
argumentative quality of a text places more weight on its grammatical
accuracy than other judges, thereby favoring texts with fewer errors but
weaker arguments. Furthermore, since the discriminal difference of the
stimuli becomes an observable outcome only through the judges'
perceptions, these biases could introduce dependencies between the
stimuli \citep{vanderLinden_et_al_2017_II}. While direct evidence for
this particular scenario is lacking, studies such as
\citet{Pollitt_et_al_2003} and \citet{vanDaal_et_al_2016} demonstrate
the presence of such biases, supporting the notion that the factors
influencing pairwise comparisons may not always cancel out.

@@

With the structure in SCM~\ref{fig-cj15_scm}, we can derive a
statistical model that addresses violations of the equal dispersion
assumption in CJ experiments (see Section~\ref{sec-theory-issue1a}).
This derivation is possible because a fully specified SCM encodes
functional and probabilistic information, which we can replace with
suitable functions and probabilistic assumptions
\citep{Pearl_et_al_2016}. Specifically, SCM~\ref{fig-cj15_scm} allows us
to express the joint distribution of our complex CJ system as a product
of simpler conditional probability distributions (CPDs)\footnote{This
  re-expression is possible because the \emph{chain rule} of probability
  and the \emph{Bayesian Network Factorization (BNF)} property. For
  further details, see \citet{Pearl_et_al_2016} and \citet{Neal_2020}.},
as shown in Equation (\ref{eq-mat51}). For clarity, we treat expressions
such as \(Y := f_{Y}(X)\), \(P(Y \mid X)\), and \(Y \sim f(Y \mid X)\)
as equivalent, where \(P(Y \mid X)\) and \(f(Y \mid X)\) represent the
CPD of \(Y\) given \(X\).
\begin{equation}\phantomsection\label{eq-mat51}{
\begin{aligned}
  P(O_{R}, & S, C, D_{R}, T_{IA}, X_{IA}, e_{IA}, T_{I}, X_{I}, e_{I}, B_{JK}, Z_{JK}, e_{JK}, B_{J}, Z_{J}, e_{J} ) & \\
  &= P(O_{R} \mid D_{R}, S, C) \cdot P(S) \cdot P(C) \cdot P(D_{R} \mid T_{IA}, B_{JK}) \\
  & \quad \cdot P(T_{IA} \mid T_{I}, X_{IA}, e_{IA}) \cdot P(T_{I} \mid X_{I}, e_{I}) \\
  & \quad \cdot P(B_{JK} \mid B_{J}, Z_{JK}, e_{JK}) \cdot P(B_{J} \mid Z_{J}, e_{J}) \\
  & \quad \cdot P(X_{IA}) \cdot P(X_{I}) \cdot P(Z_{JK}) \cdot P(Z_{J}) \\
  & \quad \cdot P(e_{IA}) \cdot P(e_{I}) \cdot P(e_{JK}) \cdot P(e_{J})  
\end{aligned}
}\end{equation}

Each CPD in Equation (\ref{eq-mat51}) rests on specific assumptions,
which we outline in SCM~\ref{fig-cj16_scm}, the probabilistic model
\ref{fig-cj16_prob}, and the statistical model \ref{fig-cj16_stat}. We
begin by assuming that \(O_{R}\) follows a Bernoulli
distribution\footnote{The binomial distribution--including its special
  case, the Bernoulli distribution--represent a maximum entropy
  distribution for binary events \citep[pp.~34]{McElreath_2020}. This
  means that the Bernoulli distribution is the most consistent
  alternative when only two un-ordered outcomes are possible and their
  expected frequencies are assumed to be constant
  \citep[pp.~310]{McElreath_2020}. For a detailed discussion of the
  binomial as a maximum entropy distribution, see
  \citet[sec.~10.1.2]{McElreath_2020}.}, reflecting the binary nature of
CJ outcomes. Furthermore, following the conventions of Generalized
Linear Models (GLMs)
\citep{Nelder_et_al_1983, Nelder_et_al_1996, Agresti_2015}, the
distribution links \(O_{R}\) to the latent discriminal difference vector
\(D_{R}\) using an inverse-logit function:
\(\text{inv\_logit}(x) = 1/(1 + \exp(-x))\).

@@

Note that while the joint distribution in Equation (\ref{eq-mat51})
includes the probability distributions of the sampling and comparison
mechanisms, \(P(S)\) and \(P(C)\), as well as those of the predictor
variables--\(P(X_{IA})\), \(P(X_{I})\), \(P(Z_{JK})\), and
\(P(Z_{J})\)--we omit these probabilities from the models presented in
Figure~\ref{fig-cj16}. We make this omission because, although these
distributions contribute to the overall joint distribution of the data,
they do not influence the relationship between \(O_{R}\) and \(D_{R}\).
This independence stems from the random selection designs that govern
\(S\), \(C\), \(X_{IA}\), \(X_{I}\), \(Z_{JK}\), and
\(Z_{J}\)\footnote{Randomization ensures that data--and, by extension,
  an estimator--satisfies several key identification properties, such as
  common support, no interference, and consistency. The most critical
  property, however, is the elimination of confounding.
  \emph{Confounding} occurs when an external variable, such as
  \(X_{I}\), simultaneously influences both the outcome (e.g.,
  \(O_{R}\)) and a variable of interest (e.g., \(S\)), resulting in
  spurious associations between the latter two
  \citep{Everitt_et_al_2010}. Randomization ensure the absence of
  confounding by effectively decoupling the association between the
  variable of interest and any other variable, except for the outcome
  itself. For a more detailed discussion on the benefits of
  randomization, see \citet{Pearl_2009}, \citet{Morgan_et_al_2014},
  \citet{Neal_2020}, and \citet{Hernan_et_al_2020}.}.

Next we define \(D_{R}\) as the difference between the discriminal
processes \(T_{IA}[i, a]\) and \(T_{IA}[h, b]\), representing the
underlying written-quality trait of the texts, plus the corresponding
repeated judge biases \(B_{JK}[j, k]\). Note that if we assume that
\(B_{JK}[j,k]\) reflects the difference in stimulus-specific biases,
i.e., \(B_{JK}[j,k] = B_{JK}[i,a,j,k] - B_{JK}[h,b,j,k]\), we can
re-write the discriminal difference as:
\begin{equation}\phantomsection\label{eq-mat52}{
\begin{aligned}
D_{R} &= \left( T_{IA}[i,a] - T_{IA}[h,b] \right) + B_{JK}[j,k] \\
&= \left( T_{IA}[i, a] + B_{JK}[i, a, j, k] \right) - \left( T_{IA}[h, b] + B_{JK}[h, b, j, k] \right) \\
& = T^{*}_{IA}[i,a] - T^{*}_{IA}[h,b]
\end{aligned}
}\end{equation}

This formulation reveals that the discriminal difference captures a
\emph{pure interaction effect}, in which neither the texts' discriminal
processes nor the judges' biases alone determine the outcome, but their
interaction does \citep{Attia_et_al_2022}. In simple terms, the
discriminal processes of the stimuli become an observable outcome only
through the judges' perceptions (biases). To clarify, the square
brackets in \(D_{R}\) indicate the relevant indices for each trait
vector; they do not imply any subsetting of the data.

@@

Next, we specify the probabilistic assumptions for the idiosyncratic
errors \(e_{I}\), \(e_{IA}\), \(e_{J}\), and \(e_{JK}\). These error
terms exhibit indeterminacies in their \emph{location},
\emph{orientation}, and \emph{scale} due to the lack of an inherent
scale in the associated latent variables. Thus, we must resolve these
indeterminacies to identify the latent variable model
\citep{Depaoli_2021, deAyala_2009}. Drawing on principles from SEM
\citep{Hoyle_et_al_2023}, we assume that the vector of idiosyncratic
errors \(\boldsymbol{e} = [e_{I}, e_{IA}, e_{J}, e_{JK}]^{T}\), follows
a Multivariate Normal distribution with mean vector \(\boldsymbol{\mu}\)
and a covariance matrix
\(\boldsymbol{\Sigma} = \boldsymbol{D} \boldsymbol{R} \boldsymbol{D}\),
with \(\boldsymbol{D}\) denoting a diagonal matrix of standard
deviations and \(\boldsymbol{R}\) a correlation matrix. To address the
\emph{location} indeterminacy, we set the errors' mean vector to zero:
\begin{equation}\phantomsection\label{eq-mat53}{
\boldsymbol{\mu} = [0, 0, 0, 0]^{T}
}\end{equation}

@@

Because the error vector \(\boldsymbol{e}\) follows an uncorrelated
Multivariate Normal distribution, the marginal distribution of
\(e_{IA}\) is a univariate Normal distribution with mean zero and
standard deviation \(p_{IA}\). We set \(P_{IA}\) as a proportion of
\(1\) to establish the scale of the stimuli' latent trait relative to
the scale of the individuals' trait. As a result, \(T_{IA}\) is also
normally distributed. This configuration effectively reinstates
Thurstone's original assumption of Normal discriminal processes for the
stimuli (see Table~\ref{tbl-thurstone_cases}).

@@

Thurstone introduced the Law of Comparative Judgment to measure
psychological traits of stimuli through pairwise comparisons
\citep{Thurstone_1927a, Thurstone_1927b}. In its general form, the
theory models single-judge comparisons across multiple, potentially
correlated stimuli. Each comparison produces a dichotomous outcome
indicating which stimulus the judge perceives as having a higher trait
level. However, Thurstone identified one key challenge in this model: it
required estimating more ``unknown'' parameters than the number of
available pairwise comparisons \citep{Thurstone_1927b}. To address this
issue and facilitate the theory's practical applicability, he formulated
five cases derived from this general form, each progressively
incorporating several simplifying assumptions.

@@

Although replacing the Normal with the Logistic distribution has a
minimal impact on trait estimation or model interpretation
\citep{vanderLinden_et_al_2017_I, McElreath_2021}, Thurstone
acknowledged that the two remaining assumptions of Case V--and, by
extension, of the BTL model--could pose significant problems when
researchers assess complex traits or heterogeneous stimuli
\citep{Thurstone_1927b}. In particular, he noted that while assuming
equal discriminal dispersions and zero correlation for the stimuli'
discriminal processes simplifies the trait measurement model, these
assumptions may fail to reflect the complexity of CJ data
\citep{Kelly_et_al_2022, Andrich_1978}. As a result, they can lead to
unreliable and inaccurate trait estimates
\citep{Ackerman_1989, Zimmerman_1994, McElreath_2020, Hoyle_et_al_2023}.

Furthermore, because Thurstone's primary goal was to produce a ``coarse
scaling'' of traits and ``allocate the compared stimuli on this
continuum'' \citep[pp.~269]{Thurstone_1927a}, his theory did not specify
how to use the resulting trait estimates for statistical inference. The
CJ tradition has attempted to address this gap by separating trait
estimation from the hypothesis testing process. Specifically, CJ studies
often rely on point estimates of the traits (i.e., the BTL scores)--or
their transformations--to conduct statistical inference. Although
separating these processes simplifies the analysis of CJ data, the
statistical literature cautions against this practice, as it can
introduce bias into the analysis and compromise the reliability of the
resulting conclusions
\citep{McElreath_2020, Kline_et_al_2023, Hoyle_et_al_2023}.

@@

The study then translates the SCM into a bespoke statistical model that
addresses key limitations of the BTL model. This model allows
researchers to analyze CJ data when the assumptions of equal dispersion
and zero correlation do not hold (see Section~\ref{sec-theory-issue1a}
and Section~\ref{sec-theory-issue1b}), and when they need to perform
statistical inference (refer to Section~\ref{sec-theory-issue2}). In
particular, the model accounts for judge biases, captures the
hierarchical structure of stimuli, accommodates heterogeneity in
discriminal dispersions, and incorporates measurement error in the
hypothesis testing process. These methodological innovations can
potentially enhance the reliability and validity of trait measurement
\citep{Perron_et_al_2015} and improve the accuracy of statistical
inferences.

@@

Five research avenues merit particular attention for their direct impact
on the reliability and validity of trait estimates and the accuracy of
statistical inferences. First, researchers should investigate how
sampling and comparison mechanisms influence CJ experimental outcomes
and affect the robustness of statistical inferences. Second, further
work is needed to understand how judges' biases impact the number of
comparisons required to achieve reliable trait estimates. Third,
researchers should re-evaluate existing methods for identifying and
excluding `misfitting' judges and stimuli to ensure their effectiveness.
Fourth, studies should explore how the hierarchical structure of stimuli
influences both experimental design and hypothesis testing in CJ
contexts. Finally, researchers should explore principled approaches for
selecting contextually relevant variables.

{About the implications of the sampling and comparison mechanisms}

{About judges biases, judgment discrepancies, and the number of judges
and judgments}

@@

{About revisit the approaches for selecting contextually relevant
variables}

Despite its potential, this study presents several challenges that
researchers must navigate. First, our approach does not establish a
clear epistemological connection between the theoretical latent trait
and its quantitative estimate. In other words, the approach does not
provide a definitive answer to the fundamental question: How do we know
that the estimated trait reflects the theoretical trait of interest?
However, this limitation is not unique to our approach; it reflects a
fundamental challenge all research involving latent constructs faces.
Nevertheless, we argue that our approach offers methodological tools
that enable researchers to design CJ experiments under conditions that
support reliable and valid trait measurement.

>>>>

Numerous studies have documented the effectiveness of CJ in assessing
traits and competencies over the past decade. These studies have
highlighted three aspects of the method's effectiveness: its
reliability, validity, and practical applicability. Research on
reliability suggests that CJ requires a relatively modest number of
pairwise comparisons \citep{Verhavert_et_al_2019, Crompvoets_et_al_2022}
to generate trait scores that are as precise and consistent as those
generated by other assessment methods
\citep{Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}. In
addition, the evidence suggests that the reliability and time efficiency
of CJ are comparable, if not superior, to those of other assessment
methods when employing adaptive comparison algorithms
\citep{Pollitt_2012b, Verhavert_et_al_2022, Mikhailiuk_et_al_2021}.
Meanwhile, research on validity indicates the capacity of CJ scores to
represent accurately the traits under measurement
\citep{Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Bartholomew_et_al_2018, Bouwer_et_al_2023}.
Lastly, research on its practical applicability highlights CJ's
versatility across both educational and non-educational contexts
\citep{Kimbell_2012, Jones_et_al_2015, Bartholomew_et_al_2018, Jones_et_al_2019, Marshall_et_al_2020, Bartholomew_et_al_2020b, Boonen_et_al_2020}.

@@

In the first scenario, research on text quality suggests that when
judges evaluate multidimensional, complex traits or heterogeneous
stimuli, they often rely on various intricate characteristics of the
stimuli to form their judgments
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}. These
additional relevant characteristics, when assessed, are unlikely to be
equally weighted or opposing. As a result, they may exert an uneven
influence on judges' perceptions, creating biases that resist
cancellation. For example, this could occur when a judge assessing the
argumentative quality of a text places more weight on its grammatical
accuracy than other judges, thereby favoring texts with fewer errors but
weaker arguments. Furthermore, since the discriminal difference of the
stimuli becomes an observable outcome only through the judges'
perceptions, these biases could introduce dependencies between the
stimuli \citep{vanderLinden_et_al_2017_II}. While direct evidence for
this particular scenario is lacking, studies such as
\citet{Pollitt_et_al_2003}, \citet{vanDaal_et_al_2016} and
\citet{Bartholomew_et_al_2020a} demonstrate the presence of such biases,
supporting the notion that the factors influencing pairwise comparisons
may not always cancel out.

@@

With the structure in SCM~\ref{fig-cj15_scm}, we can derive a
statistical model that addresses violations of the equal dispersion
assumption (see Section~\ref{sec-theory-issue1a}). This derivation is
possible because a fully specified SCM encodes functional and
probabilistic information, which we can replace with suitable functions
and probabilistic assumptions \citep{Pearl_et_al_2016}. Specifically,
SCM~\ref{fig-cj15_scm} allows us to express the joint distribution of
our complex CJ system as a product of simpler conditional probability
distributions (CPDs)\footnote{This re-expression is possible because the
  \emph{chain rule} of probability and the \emph{Bayesian Network
  Factorization (BNF)} property. For further details, see
  \citet{Pearl_et_al_2016} and \citet{Neal_2020}.}, as shown in Equation
(\ref{eq-mat51}). For clarity, we treat expressions such as
\(Y := f_{Y}(X)\), \(P(Y \mid X)\), and \(Y \sim f(Y \mid X)\) as
equivalent, where \(P(Y \mid X)\) and \(f(Y \mid X)\) represent the CPD
of \(Y\) given \(X\). \begin{equation}\phantomsection\label{eq-mat51}{
\begin{aligned}
  P(O_{R}, & S, C, D_{R}, T_{IA}, X_{IA}, e_{IA}, T_{I}, X_{I}, e_{I}, B_{JK}, Z_{JK}, e_{JK}, B_{J}, Z_{J}, e_{J} ) & \\
  &= P(O_{R} \mid D_{R}, S, C) \cdot P(S) \cdot P(C) \cdot P(D_{R} \mid T_{IA}, B_{JK}) \\
  & \quad \cdot P(T_{IA} \mid T_{I}, X_{IA}, e_{IA}) \cdot P(T_{I} \mid X_{I}, e_{I}) \\
  & \quad \cdot P(B_{JK} \mid B_{J}, Z_{JK}, e_{JK}) \cdot P(B_{J} \mid Z_{J}, e_{J}) \\
  & \quad \cdot P(X_{IA}) \cdot P(X_{I}) \cdot P(Z_{JK}) \cdot P(Z_{J}) \\
  & \quad \cdot P(e_{IA}) \cdot P(e_{I}) \cdot P(e_{JK}) \cdot P(e_{J})  
\end{aligned}
}\end{equation}

Each CPD in Equation (\ref{eq-mat51}) rests on specific assumptions,
which we outline in Figure~\ref{fig-cj16}. We begin by assuming that
\(O_{R}\) follows a Bernoulli distribution\footnote{The binomial
  distribution--including its special case, the Bernoulli
  distribution--represent a maximum entropy distribution for binary
  events \citep[pp.~34]{McElreath_2020}. This means that the Bernoulli
  distribution is the most consistent alternative when only two
  un-ordered outcomes are possible and their expected frequencies are
  assumed to be constant \citep[pp.~310]{McElreath_2020}. For a detailed
  discussion of the binomial as a maximum entropy distribution, see
  \citet[sec.~10.1.2]{McElreath_2020}.}, reflecting the binary nature of
CJ outcomes. Furthermore, following the conventions of Generalized
Linear Models (GLMs)
\citep{Nelder_et_al_1983, Nelder_et_al_1996, Agresti_2015}, the
distribution links \(O_{R}\) to the latent discriminal difference vector
\(D_{R}\) using an inverse-logit function:
\(\text{inv\_logit}(x) = 1/(1 + \exp(-x))\).

@@

Note that while the joint distribution in Equation (\ref{eq-mat51})
includes the probability distributions of the sampling and comparison
mechanisms, \(P(S)\) and \(P(C)\), as well as those of the predictor
variables--\(P(X_{IA})\), \(P(X_{I})\), \(P(Z_{JK})\), and
\(P(Z_{J})\)--we omit these probabilities from the models presented in
Figure~\ref{fig-cj16}. We make this omission because, although these
distributions contribute to the overall joint distribution of the data,
the variables \(S\), \(C\), \(X_{IA}\), \(X_{I}\), \(Z_{JK}\), and
\(Z_{J}\) are observed and independent of any other variable in the
model. As observed variables, they do not require distributional
assumptions--unlike the idiosyncratic errors. Moreover, their
independence stems from the random selection designs that govern the
variables\footnote{Randomization ensures that data--and, by extension,
  an estimator--satisfies several key identification properties, such as
  common support, no interference, and consistency. The most critical
  property, however, is the elimination of confounding.
  \emph{Confounding} occurs when an external variable, such as
  \(X_{I}\), simultaneously influences both the outcome (e.g.,
  \(O_{R}\)) and a variable of interest (e.g., \(S\)), resulting in
  spurious associations between the latter two
  \citep{Everitt_et_al_2010}. Randomization ensure the absence of
  confounding by effectively decoupling the association between the
  variable of interest and any other variable, except for the outcome
  itself. For a more detailed discussion on the benefits of
  randomization, see \citet{Pearl_2009}, \citet{Morgan_et_al_2014},
  \citet{Neal_2020}, and \citet{Hernan_et_al_2020}.}.

Next we define \(D_{R}\) as the difference between the discriminal
processes \(T_{IA}[i, a]\) and \(T_{IA}[h, b]\), representing the
underlying written-quality trait of the compared texts, plus the
corresponding repeated judge bias \(B_{JK}[j, k]\). Note that if we
assume that \(B_{JK}[j,k]\) reflects the difference in stimulus-specific
biases, i.e., \(B_{JK}[j,k] = B_{JK}[i,a,j,k] - B_{JK}[h,b,j,k]\), we
can re-write the discriminal difference as:
\begin{equation}\phantomsection\label{eq-mat52}{
\begin{aligned}
D_{R} &= \left( T_{IA}[i,a] - T_{IA}[h,b] \right) + B_{JK}[j,k] \\
&= \left( T_{IA}[i, a] + B_{JK}[i, a, j, k] \right) - \left( T_{IA}[h, b] + B_{JK}[h, b, j, k] \right) \\
& = T^{*}_{IA}[i,a] - T^{*}_{IA}[h,b]
\end{aligned}
}\end{equation}

This formulation reveals that the discriminal difference captures a
\emph{pure interaction effect}, in which neither the texts' discriminal
processes nor the judges' biases alone determine the outcome, but their
interaction does \citep{Attia_et_al_2022}. Put simply, this mathematical
description captures the idea that the stimuli' discriminal processes
become an observable outcome only through the lens of judges'
perceptions (i.e., their biases). For clarity, the square brackets in
\(D_{R}\) indicate the relevant indices for each trait vector; they do
not imply any subsetting of the data.

@@

Next, we specify the probabilistic assumptions for the idiosyncratic
errors \(e_{I}\), \(e_{IA}\), \(e_{J}\), and \(e_{JK}\). Unlike other
variables in the model, these error terms exhibit indeterminacies in
their \emph{location}, \emph{orientation}, and \emph{scale} due to the
lack of an inherent scale in the associated latent variables \(T_{I}\),
\(T_{IA}\), \(B_{J}\), and \(B_{JK}\). Thus, to identify the latent
variable model, we must resolve these indeterminacies
\citep{Depaoli_2021, deAyala_2009}. Drawing on principles from SEM
\citep{Hoyle_et_al_2023}, we assume that the vector of idiosyncratic
errors \(\boldsymbol{e} = [e_{I}, e_{IA}, e_{J}, e_{JK}]^{T}\), follows
a Multivariate Normal distribution with mean vector \(\boldsymbol{\mu}\)
and a covariance matrix
\(\boldsymbol{\Sigma} = \boldsymbol{D} \boldsymbol{R} \boldsymbol{D}\),
with \(\boldsymbol{D}\) denoting a diagonal matrix of standard
deviations and \(\boldsymbol{R}\) a correlation matrix. To address the
\emph{location} indeterminacy, we set the errors' mean vector to zero:
\begin{equation}\phantomsection\label{eq-mat53}{
\boldsymbol{\mu} = [0, 0, 0, 0]^{T}
}\end{equation}

@@

Note that because the error vector \(\boldsymbol{e}\) follows an
uncorrelated Multivariate Normal distribution, the marginal distribution
of \(e_{IA}\) is a univariate Normal distribution with mean zero and
standard deviation \(p_{IA}\). We set \(P_{IA}\) as a proportion of
\(1\) to establish the scale of the stimuli' latent trait relative to
the scale of the individuals' trait. As a result, \(T_{IA}\) is also
normally distributed. This configuration effectively reinstates
Thurstone's original assumption of Normal discriminal processes for the
stimuli (see Table~\ref{tbl-thurstone_cases}).

@@

Thurstone introduced the Law of Comparative Judgment to measure
psychological traits of stimuli through pairwise comparisons
\citep{Thurstone_1927a, Thurstone_1927b}. In its general form, the
theory models single-judge comparisons across multiple, potentially
correlated stimuli. Each comparison produces a dichotomous outcome
indicating which stimulus the judge perceives as having a higher trait
level. However, Thurstone identified one key challenge with its
measurement model: it required estimating more ``unknown'' parameters
than the number of available pairwise comparisons
\citep{Thurstone_1927b}. To address this issue and facilitate the
theory's practical applicability, he formulated five cases derived from
this general form, each progressively incorporating several simplifying
assumptions.

@@

Although replacing the Normal with the Logistic distribution has a
minimal impact on trait estimation or model interpretation
\citep{vanderLinden_et_al_2017_I, McElreath_2021}, Thurstone
acknowledged that the two remaining assumptions of Case V--and, by
extension, of the BTL model--could pose significant problems when
researchers assess complex traits or heterogeneous stimuli
\citep{Thurstone_1927b}. In particular, he noted that while assuming
equal discriminal dispersions and zero correlation for discriminal
processes simplifies the trait measurement model, these assumptions may
fail to reflect the complexity of CJ data
\citep{Thurstone_1927a, Andrich_1978, Kelly_et_al_2022}. As a result,
they can lead to unreliable and inaccurate trait estimates
\citep{Ackerman_1989, Zimmerman_1994, McElreath_2020, Hoyle_et_al_2023}.

Furthermore, because Thurstone aimed primarily to produce a trait's
``coarse scaling'' and ``allocate the compared stimuli on this
continuum'' \citep[pp.~269]{Thurstone_1927a}, his theory did not specify
how to use the resulting trait estimates for statistical inference. The
CJ tradition has attempted to address this gap by separating trait
estimation from the hypothesis testing process. Specifically, CJ studies
often rely on point estimates of the traits (i.e., the BTL scores)--or
their transformations--to conduct statistical inference. Although
separating these processes simplifies the analysis of CJ data, the
statistical literature cautions against this practice, as it can
introduce bias into the analysis and compromise the reliability of the
hypothesis test
\citep{McElreath_2020, Kline_et_al_2023, Hoyle_et_al_2023}.

@@

The study then translates the SCM into a bespoke statistical model that
addresses key limitations of Case V. This model allows researchers to
analyze CJ data when the assumptions of equal dispersion and zero
correlation do not hold (see Section~\ref{sec-theory-issue1a} and
Section~\ref{sec-theory-issue1b}), and when they need to perform
statistical inference (refer to Section~\ref{sec-theory-issue2}). In
particular, the model accounts for judge biases (see
Section~\ref{sec-theory-theoretical_P1}), captures the hierarchical
structure of stimuli (refer to Section~\ref{sec-theory-theoretical_P3}),
incorporates measurement error in the hypothesis testing process (refer
to Section~\ref{sec-theory-theoretical_P3}), and accommodates
heterogeneity in discriminal dispersions (see
Section~\ref{sec-statistical}). These methodological innovations can
potentially enhance the reliability and validity of trait measurement
\citep{Perron_et_al_2015} and improve the accuracy of statistical
inferences.

@@

Five research avenues deserve particular attention due to their direct
impact on the reliability and validity of trait estimates and the
accuracy of statistical inferences. First, researchers should
investigate how sampling and comparison mechanisms influence CJ
experimental outcomes and affect inference robustness. Second, studies
should assess the impact of judges' bias on the number of comparisons
needed to achieve reliable trait estimates. Third, future work should
re-evaluate current methods for identifying `misfitting' judges and
stimuli. Fourth, researchers should analyze how the hierarchical
structure of stimuli affects both experimental design and hypothesis
testing. Finally, studies should explore how modeling non-dichotomous
outcomes or incorporating multidimensional traits affects the
reliability and validity of the trait estimates.

The explicit integration of the sampling and comparison mechanisms
within our conceptual model offers a novel perspective on the first
research avenue. Specifically, it shows how these mechanisms influence
the data-generating process by acting as missing data sources,
determining which observations are part of the data sample. This new
perspective invites the application of Little and Rubin's principled
missing data framework \citep{Little_et_al_2020}, enabling a more
rigorous evaluation of existing claims about these mechanisms, their
influence on CJ outcomes, and their implications for designing and
evaluating more complex experimental setups.

One prominent claim in the CJ literature is that Thurstone's model
naturally handles non-random missing data without compromising the
reliability or validity of trait estimates \citep{Bramley_2008}.
However, it is easy to imagine scenarios where this assumption may not
hold. For instance, selecting a particular group of judges could result
in an incomplete representation of the trait's full range and
complexity, as judgment accuracy varies with demographic factors
\citep{Bartholomew_et_al_2020a, Kelly_et_al_2022} and even between
individuals
\citep{Gill_et_al_2013, vanDaal_et_al_2016, vanDaal_et_al_2017, vanDaal_2020}.
Consequently, judges' selection could affect the formation of a ``shared
consensus'', potentially biasing trait estimates and distorting the
observed trait distribution \citep{Deffner_et_al_2022}. It is therefore
essential to determine whether such sampling biases occur--and if so,
whether they undermine the validity of trait estimates and the
robustness of statistical inferences. Our framework offers a rigorous
setting to test the scenario.

{another example: For instance, consider a CJ experiment assessing
speech quality using a case-control design \citep{Everitt_et_al_2010}
that samples individuals with normal hearing (NH) and those with hearing
impairment (HI). Here, the sampling mechanism depends on hearing
status--a variable likely to influence the trait of interest, as HI
individuals may exhibit lower speech quality
\citep{Boonen_et_al_2023, Rivera_et_al_2024}. Suppose a subgroup within
the HI population, such as individuals with genetically induced hearing
loss, is difficult to reach. If this group is systematically excluded
and happens to over- or under-perform in speech quality, the resulting
trait estimates may be biased. It is therefore essential to determine
whether such biases occur--and if so, whether they undermine the
validity of trait estimates and the robustness of statistical
inferences--and our framework provides a rigorous setting to test this
claim.}

{another example: For instance, consider a high-stakes non-mandatory
summative CJ assessment, where the results the individuals have in this
assessment have some implications for them, however, their participation
is not mandatory. In this case, it not hard to imagine that some of
these individuals will not participate in the CJ assessment, thus
affecting the sample, and ultimately the trait estimates: now a probable
under performing group will not available in the data sample, thus
biasing the resulting trait estimates. It is therefore essential to
determine whether such biases occur--and if so, whether they undermine
the validity of trait estimates and the robustness of statistical
inferences--and our framework provides a rigorous setting to test this
claim.}

Moreover, when designing and evaluating more complex experimental
setups, our framework highlights one key principle: the sampling and
comparison mechanisms directly influence the CJ outcome. In this study,
we deliberately structured these mechanisms to produce data that is
\emph{missing completely at random} (MCAR)--that is, the missingness is
unrelated to any observed or unobserved variables, including the outcome
\citep{Little_et_al_2020}. This design offers one key advantage: it
generates simple random samples that satisfy the condition of
ignorability, enabling researchers to legitimately ignore missing data
during analysis without introducing bias
\citep{Everitt_et_al_2010, Kohler_et_al_2019}.

However, many modern CJ applications involve more complex forms of
missingness--such as \emph{missing at random} (MAR) or \emph{missing not
at random} (MNAR) mechanism \citep{Little_et_al_2020}--that arise from
the chosen sampling and comparison methods. Two notable examples are
adaptive comparative judgment designs
\citep{Pollitt_2012a, Pollitt_2012b} and case-control studies
\citep{Everitt_et_al_2010}. In adaptive designs, an algorithm defines
the comparison mechanism by dynamically selecting which stimulus pairs
judges evaluate. These designs have produced mixed evidence: some
studies report improved trait reliability
\citep{Pollitt_et_al_2003, Pollitt_2012a, Pollitt_2012b}, while others
suggest these gains may be artificially inflated
\citep{Bramley_2015, Bramley_et_al_2019, Crompvoets_et_al_2020, Crompvoets_et_al_2022}.
In contrast, case-control studies rely on sampling mechanisms linked to
non-assignable variables that likely influence the trait of interest.
For instance, \citet{Boonen_et_al_2020} selected children based on
hearing status, a non-assignable variable that likely affects speech
quality \citep{Boonen_et_al_2023, Rivera_et_al_2024}. Nevertheless,
regardless of their missingness mechanism, both designs could benefit
from explicitly articulating their underlying causal structures. Doing
so would enable researchers to assess and test their impact on trait
estimation and statistical inference. Our framework provides a rigorous
foundation for such articulations.

Additionally, by explicitly incorporating judges' biases, our framework
provides a novel approach to investigate the second research avenue.
Specifically, it enables researchers to determine whether such biases
occur and, if so, to assess their impact on the reliability and validity
of trait estimation. Although some anecdotal evidence already suggests
judges' biases exist
\citep{Pollitt_et_al_2003, vanDaal_et_al_2016, Bartholomew_et_al_2020a},
further research is needed to address several key questions. For
instance: Can CJ data be analyzed under the assumption of
``sample-free'' trait calibration--that is, the estimated trait
difference for any two scripts does not depend on which other scripts
they are compared with, implying an absence of judges' bias?
\citep{Bramley_2008, Andrich_1978} If this assumption does not hold, how
many judges and judgments are required to ensure acceptable levels of
trait reliability and validity? Do training and expertise influence
judgment biases, and to what extent? \citep{Kelly_et_al_2022} Finally,
what are the benefits of \emph{repeated measures designs}, in which
judges evaluate the same stimulus pairs multiple times?
\citep{Lawson_2015}

@@

{About non-dichotomous outcomes and multidimensional traits}

While a natural step would be to formally derive answers to all these
research avenues--similar to approaches used in classical models like
linear regression--we argue that the complexity of the CJ system makes
this route impractical. Instead, simulation-based methods, like power
analysis, offer a more feasible alternative, allowing researchers to
investigate these issues without relying on cumbersome mathematical
proofs.

Despite its potential, this study presents several challenges that
researchers must navigate. First, researchers need to be proficient in
Probabilistic Programming Languages (PPLs) and Bayesian inference
methods. This expertise is essential for translating context-specific
conceptual models into bespoke statistical models that meaningfully
address the research questions at hand. While such translations are not
exclusive to Bayesian methods, we argue that Bayesian approaches offer a
more accessible path. They reduce, for instance, the need for additional
expertise in optimization techniques and theory often required by
frequentist methods, thereby lowering the technical barrier for many
applied researchers.
